---
title: 'Duq Biostats Week 6 | Regression Tutorial 1: Data Cleaning'
author: "brouwern@gmail.com"
date: "February 12, 2017"
output:
  html_document:
    toc: yes
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this tutorial we'll 

* load data from Sibiel et al 2013 in .csv format
* fix some typos  
* clean up some columns of numeric data that contain charcter data
* save the data as a .csv for future use

## Important functions used

* setwd
* read.csv
* grep
* gsub
* as.numeric
* which
* write.csv()


## Original Data

**Skibiel et al 2013.**  Journal of Animal Ecology.  The evolution of the nutrient composition of mammalian milks. 82: 1254-1264.



# Load data

## Set working directory

* "working directory" = The location of where I have the data.
* Not the same as the "workspace"
* Use RStudio to set your wd
    + Click on "Session"
    + "Set working directory"
    + "Choose directory"
* Code to get the working directory will appear in the console
* Copy and paste this into a code chunk for easy future access


For my computer, this is the full "path" to my current R working directory.
```{r}
setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/1_STAT_Duq_2017/HOMEWORK/Duq_Biostats_Homework_4_Milk_regression/homework4_RMD_vs2")
```

## Load data from csv

I have cleaned the data and set it up in a csv file saved into the working directory.
```{r}
dat <- read.csv(file = "Skibiel_clean_milk.csv")
```

### Look at data

```{r, echo = F}
head(dat)
tail(dat)
dim(dat)
summary(dat)
```

<br>

## Fix NAs

After working with the data a bit I realized I had 2 NA values in my "fat" column that should't be there.  I use the following code to

* Identify the location of the NAs with is.na()
* Identify the specific row of data by using which() to ID the specific species with an NA for fat
* Overwrite the NA with the correct data using row indexing
    + ie "dat[i.H.niger,"fat"] <- 5.0"
* Note that is.na() returns TRUE/FALSE values; "TRUE" acts like specific index when you use it for row indexing
    + ie, even though "i.NA" is a vector of TRUES and FALSE, dat$spp[i.NA]" returns just the two species that match the "TRUES" 
Normally I would make these changes in my raw data file, but its useful to see how these changes can be made within R.


### Fix NAs by hand

```{r}
# Look at fat column
summary(dat$fat)

# Find NAS
i.NA <- is.na(dat$fat)

#Identify the species w/ NAs
dat$spp[i.NA]

# Extract the indexes of the appropriate species
i.H.niger <- which(dat$spp == "Hippotragus niger")
i.C.elaphus <- which(dat$spp == "Cervus elaphus hispanicus")

#Overwrite the bad values
dat[i.H.niger,"fat"] <- 5.0
dat[i.C.elaphus,"fat"] <- 12.6


```



# Clean variables

* There are some annotations from the raw data still hanging out in the dataframe
* I will use "regular expressions" to clean these up
* Regular expression take practice to use but there are many resources online for learning about them
    + cleaning up asterisks (*), as below, is even a little tricker.
* What follows is just a brief snapshot into this feature of R
    + R is known, however, for not having stellar basic regular expression features.


## Clean Protein column

There are letters, commas, and asterisks that were annotations in the original datafile.  Easy to remove with find-replace, but also very easy with gsub(), with not chance of changing anything else in your spreadsheet you dont want

Look at all of the protein column
```{r}
summary(dat$protein)
```

### grep()

* Use grep() to find the index of things
* "[a-zA-Z]" = "find all letter, lower and uppercase"
```{r}
i.letters <- grep("[a-zA-Z]",dat$protein)

dat$protein[i.letters]
```

There's jsut an "S". I coudl find it like this with grep too

```{r}
#these are equivalent for the simple case of a single letter
i.S <- grep("S",dat$protein)  #no brackets
i.S <- grep("[S]",dat$protein)#brackets

dat$protein[i.S]

```

### Use the gsub() command to replace letter

```{r}
dat$protein <- gsub("[a-zA-Z]", #pattern
                    "",         #replace
                    dat$protein)#data
```

Check
```{r}
grep("[a-zA-Z]",dat$protein)
```


### Use the gsub() command to replace asterisks *

* Asterisks are a "special character" for regular expressions
*These can be tricky
* Have to put \\ in front of it

```{r}
dat$protein <- gsub("\\*","",dat$protein)
```

Check
```{r}
grep("\\*",dat$protein)
```

### Use the gsub() command to replace comma

Commas are also species characters

```{r}
dat$protein <- gsub("\\,","",dat$protein)
```

Check
```{r}
grep("\\,",dat$protein)
```


### Check for space

These are frequent typos in data

```{r}
grep(" ", dat$protein)
```
 None
 
 
## Convert character data to numeric data

* Because there were non-numeric characters (S, a comamn, *) in the column, are loaded "protein" as character data; basicly it treated it as words/symbols but not numbers.
* We now need to convert it back to numbers
* This is done with the command as.numeric()
* Converting back and forth can somtimes cause problems, so its good to put the converted info in a new column to check against the old

 
```{r}
dat$protein.num <- as.numeric(dat$protein)
```

Compare old and new columns

```{r}
head(dat[,c("protein.num","protein")])
tail(dat[,c("protein.num","protein")])
```

## Clean other columns

Check other columns
```{r}
summary(dat[,c("sugar","energy")])
```

### Sugar

* The sugar column has "<" and letters.  
* I can put anything I want removed inside the brackets
    + "[a-zA-Z<]"" wiil get rid of the letters and "<"

```{r}
dat$sugar <- gsub("[a-zA-Z<]","",dat$sugar)
```

Convert and check
```{r}
dat$sugar.num <- as.numeric(dat$sugar)

head(dat[,c("sugar.num","sugar")])

```



## Energy

```{r}
#The followign are equivalent
summary(dat[,c("energy")])
summary(dat$energy)

#clean
dat$energy <- gsub("[a-zA-Z<]","",dat$energy)

#convert and check
dat$energy.num <- as.numeric(dat$energy)

head(dat[,c("energy.num","energy")])


```



# Save the cleaned data

```{r}
write.csv(dat, "Skibiel_clean2_milk.csv")
```

