---
title: "Duq Biostats Week 6 | Regression Tutorial 5: Functions"
author: "brouwern@gmail.com"
date: "February 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions in R

A simple function to calculate the standard SE

## Standard error

### Equation
\[SE = \frac{SD}{\sqrt{N}}\]

<br>

### R-ish notation
\[SE = \frac{SD(x)}{sqrt(length(x))}\]




## Calcualte SE by hand

Make some fake data
```{r}
fake.data <- rnorm(n = 100,mean = 0,sd = 1)
```

<br>

Calc SE
```{r}
sd(fake.data)/sqrt(length(fake.data))
```

<br>

# A function that does these calculation

First I'll make a function that breaks the steps up 
```{r}

SE.function <- function(x){ 
      
      #Calcualte the SD
      sd.x <- sd(x)
      
      #sample size
      length.x <- length(x)
      
      #sqrt of sample size
      sqrt.length <- sqrt(length.x)
      
      #calculate the SE
      SE <- sd.x/sqrt.length
      
      #return teh S
      return(SE)
       }



```

```{r}
SE.function(fake.data)
```

<br><br>

This function does the same thing directly

```{r}

SE.function <- function(x){ sd(x)/sqrt(length(x)) }

```


<br><br>

## Test the function


```{r}

(fake.data)

```

<br>

Compare to a real function
```{r}

library(plotrix)

```

<br>

Check 
```{r}
std.error(fake.data)
```

How exact is my function?

```{r}
std.error(fake.data) == SE.function(fake.data)
```

<br><br>

# Problem: My function will choke if there's an NA.


Make the first entry in the vector NA
```{r}
fake.data[1] <- NA
```

<br>

Check
```{r}
std.error(fake.data)
SE.function(fake.data)
```

<br>


This is b/c the sd function doesnt like NAs
```{r}
sd(fake.data)
```

<br>

I'll fix the function by add "na.rm = TRUE" to the function

```{r}

SE.function <- function(x){sd(x, na.rm = T)/sqrt(length(x))}

```

<br><br>

Now it doesn't choke, but it gives the wrong asnwer

```{r}
#correct
std.error(fake.data)

#mine
SE.function(fake.data)
```

<br>

This is b/c the length() data gives the totally length of the vector, including any NAs

```{r}
length(fake.data)
```

<br>

length() doesn't have an arugement related to NAs.  So I need to remove them first before I do the calculations.  I'll use the na.omit() function to do this

```{r}
SE.function2 <- function(x){
  # remove NAs
  x2 <- na.omit(x)
  
  # calcualte
  sd(x2)/sqrt(length(x2))}
```

<br>


Check the results
```{r}
std.error(fake.data)
SE.function2(fake.data)
```

<br><br><br>

# A more complex function

Writing lots of plotting code can be a drag in R.  I'll make a function that can speed this up.

## Regression model

Normally to plot a regression line we'd do something like this

Load milk data
```{r}
setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/1_STAT_Duq_2017/HOMEWORK/Duq_Biostats_Homework_4_Milk_regression/homework4_RMD_vs2")
dat2 <- read.csv("milk_subset.csv")
```

<br>

```{r}
#the the model
model. <-lm(fat.log10 ~ mass.log10, data = dat2)

#plot the raw dat
plot(fat.log10 ~ mass.log10, data = dat2)

#plot regression line
abline(model., col = 2)

```


<br><br>

We can make a function the simplifies this a bit
```{r}
plot.my.mod <- function(mod){
  
  #extract formula from model
  model.formula <- formula(mod)
  
  #plot the raw dat
  plot( model.formula , data = dat2)
  
  #plot regression line
  abline(mod, col = 2)

}
```


<br>

Test it
```{r}
plot.my.mod(model.)
```

<br>


I'd like to be able to change the colors of the regression line on the fly

* add "..." to the function() code

```{r}
plot.my.mod <- function(mod, ...){
  
  #extract formula from model
  model.formula <- formula(mod)
  
  #plot the raw dat
  plot( model.formula , data = dat2)
  
  #plot regression line
  abline(mod, ...)

}
```

<br><br>

Test it
```{r}
plot.my.mod(model., col = 3)
```


<br>

I can change the line type of the regression line now too
```{r}
plot.my.mod(model., col = 3, lty = 2)
```

<br>

# A more complex situation


* Build a model with a continuous and categorical predictor.  
* This model defines two parralell lines, one for aquatic species, one for terrestrial species.

Build the model
```{r}
model.2 <-lm(fat.log10 ~ mass.log10 + 
               biome, 
             data = dat2)


summary(model.2)
```


**Question**
What are the equations for these two lines?


<br><br><br>

My function won't work properly on model output this b/c it uses the abline() function, which does't work w/ complex regression models

```{r}
plot.my.mod(model.2, col = 3, lty = 2)
```

<br>


## A function to plot models with two predictors

In a files called "fnxn_plot_intxns.R"  I have a function that can plot this model properly.  I will load this function using source()

```{r}
source("fnxn_plot_ANCOVA.R")
```

<br>

* The function is called plot.ANCOVA()
* As written this function will not work if you did the log transformations "on the fly!"
    * MUSt have transformation "hard coded" as a seperate column

```{r}
plot.ANCOVA(model.2)
```




<br><br><br><br><br><br><br><br><br><br><br><br><br>

# Code for the plot.ANCOVA()

```{r}


#makes interaction plot of simple regression
#models w/1 numeric and 1 categorical predictor



#makes interaction plot of simple regression
#models w/1 numeric and 1 categorical predictor


plot.ANCOVA <- function(model,
                        raw.data = NULL,
                        x.axis = NULL,
                        cols = c(1,2)){
  
  #Data
  ##extract data used in model
  dat <- model$model
  
  ##if x.axis arguement contains data, add to dataframe
  if(is.character(x.axis) == TRUE){
    dim.orig <- dim(dat)[2]
    dat <- cbind(dat,raw.data[,x.axis])
    names(dat)[-c(1:dim.orig)] <-  x.axis
  }
  
  
  
  #extract formula from model object
  ##get formula
  form <- as.character(formula(model))[-1]
  
  ##split into vector
  model.terms <- strsplit(form[2],"[\\+\\*]")
  
  ##name list returned by strsplit()
  names(model.terms) <- "x.names"
  
  ##get rid of spaces
  model.terms$x.names <- gsub(" ","" ,model.terms$x.names)
  
  ##count number of parameters
  i.terms <- length(model.terms$x.names)
  
  #create new entry in list (this will be overwritten)
  model.terms$x.types <- model.terms$x.names
  
  #Loop over each term in the model to determine
  ##whether it is categorical or continous
  for(i in 1:i.terms){
    term.i <- model.terms$x.names[i]
    
    if(any(strsplit(term.i,"")[[1]] == "*") == TRUE){
      model.terms$x.types[i] <- "interactions"
      next
    }
    
    
    model.terms$x.types[i] <- ifelse(is.numeric(dat[,term.i]) == TRUE, 
                                     "numeric",
                                     "categorical")
  }
  
  
  
  ## y axis
  model.terms$y <- form[1]
  
  
  
 
  
  #Define continous variable
  ## If the model doesn't contain a continuous, 
  ## get it from x.axis arguement
  if(is.null(x.axis)== FALSE){
    x.cont <- x.axis
  }
  
  ##If model contains a continous variable...
  if(is.null(x.axis)== TRUE){
    i.num <- which(model.terms$x.types == "numeric")
    x.cont <- model.terms$x.names[i.num]
    
  }

  
  #Define categorical variable
  i.cat <- which(model.terms$x.types == "categorical")
  x.cat <- model.terms$x.names[i.cat]
  
  if(length(x.cat) > 1){
    x.cat <- x.cat[1]
    
  }
  
  
  #Define y variable
  y <- model.terms$y
  
  

  
  # range of observed numeric data
  ## if calling from dataframe
  if(is.character(x.cont) == TRUE){
    rng <- range(dat[,x.cont])
    
  }
  
  ## if calling from x.var arguement
  if(is.numeric(x.cont) == TRUE){
    rng <- range(x.cont)
    
  }
  
  ## levels of the categorical variable
  levs <- levels(dat[,x.cat])
  
  
  ## generate new data with expand.grid() 
  newdat <- expand.grid(x.cont = rng,
                        x.cat = levs)
  

  ## label columns

  #if calling from x.axis
  if(is.numeric(x.cont) == TRUE){
    names(newdat) <- c("x.axis", x.cat)
    
  }
 
  #if calling from model datat
  if(is.character(x.cont) == TRUE){
    names(newdat) <- c(x.cont, x.cat)
    
  }
  
  # generate predictions from fitted model
  newdat$yhat <- predict(model,newdat)
  
  #determine baseline level of fitted model
  i.lev1 <- which(dat[,x.cat] == levs[1])
  
  #determine range of values
  ## range of observed x values
  #if calling from x.axis
  if(is.numeric(x.cont) == TRUE){
    xlims <- range(x.cont)
    
    
  }
  
  #if calling from model datat
  if(is.character(x.cont) == TRUE){
    xlims <- range(dat[,x.cont])
    
    
  }
  
 
  
  
  
  ## range of observed y values
  ylims <- range(dat[,y])
  
  
  #Plot raw data
  ## Plot level 1 of model (intercept)
  plot(dat[i.lev1,y] ~ dat[i.lev1,x.cont], 
       xlab = x.cont,
       ylab = y,
       xlim = xlims,
       ylim = ylims,col=cols[1])
  
  ## Plot level 2 of model (intercept+effect)
  points(dat[-i.lev1,y] ~ dat[-i.lev1,x.cont], 
         xlab = x.cont,
         ylab = y,
         col = cols[2])
  
  
  #Plot regession lines
  ##subset data for regression lines
  i.lev1 <- newdat[,x.cat] == levs[1]
  i.lev2 <- newdat[,x.cat] == levs[2]
  
  ##plot regression lines
  points(newdat$yhat ~ newdat[,x.cont], 
         subset = i.lev1,
         col = cols[1],
         type = "l")
  
  points(newdat$yhat ~ newdat[,x.cont], 
         subset = i.lev2,
         col = cols[2],
         type = "l")
  
  
}





```

