---
title: "Regression Data Exploration: Skibiel et al 2013 milk data"
author: "brouwern@gmail.com"
date: "February 1, 2017"
output:
  word_document:
    reference_docx: rmarkdown_template.docx
    fig_width: 9
    fig_height: 5
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = TRUE)
```



# Introduction to Regression Modeling

<br>

# Data for today: comparative zology of milk
## How does milk compositon vary with animal life history & biology?

**Skibiel et al 2013.**  Journal of Animal Ecology.  The evolution of the nutrient composition of mammalian milks.

* Table S1. Milk composition of mammals at mid-lactation
* Table S2. Ecology and life histories of mammals whose milk composition has been described

<br>

# The milk dataset

* Focus on % of fat in milk
* Several "continous predictors" / covariates
    + mass of female = mass.female 
    + duration of pregnancy/gestation = gest.month.NUM    
    + how long young fed = lacat.mo.NUM     
    + how big are babies = mass.litter         
* Several "categorical predictors"
    + diet = meat, plants, everything
    + arid = desert or other
    + biome = aquatic or terrestrial (includes bats)

```{r, echo = F, display = F}
#Data cleaned in 1_REGRESSIOn_data_cleaning_Skibiel_et_al...
milk2 <- read.csv(file = "./Skibiel_clean_milk_focal_column.csv")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Look at data
    
```{r}
# Hom much data?
dim(milk2)

#focal columns
##(use negative indexing to drop some)
summary(milk2[,-c(2,3,8,9)])
```



## Focal families in dataset

* Artiodactyla = 23; Even-toed ungulate
* Carnivora = 23
* Primates = 22
* Rodentia = 17
* Chiroptera = 10; bats
* Diprotodontia = 10; marsupials
* Cetacea = 6
* Perrissodactyla = 7; odd-toed ungulates
* Lagomorpha = 3

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>




# Explore milk data with ggplot2

* ggplot2::qplot is an excellent function for rapid data exploration
* different synatax than R's "base" plot functions
* (another R plotting system is "trellis")

## Look at % milk fat vs. mass of female

* Plot is dominated by whales!
```{r gg1, echo = F}
library(ggplot2)

#set font
theme_set(theme_bw(base_size = 18))

qplot(y = fat.percent,
      x = mass.female,
      data = milk2,
      main = "Regression data: continous vs continuous",
      xlab = "Continous x: mass of female",
      ylab = "Continous y: mass of female") 
```

```{r gg1, eval = F}
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Look at % milk fat vs. log(mass of female)

* Logging helps
* NOTE: statiticians typically use natural log = ln()
* Biologists, especially when allometry is invovled, typically use base 10 = log10()
* Original authors used log10, but I just realized this, so I'll stick with ln()
* I believe (?) give similar answers, but log10 I think (?) makes biological interpretation easier (???)

```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = log(mass.female),
      data = milk2,
            main = "Regression data: continous vs log(continuous)",
      xlab = "Continous x: log(mass of female)",
      ylab = "Continous y: mass of female")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Compare log10

* Looks very similar

```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = log10(mass.female),
      data = milk2,
            main = "Regression data: continous vs log(continuous)",
      xlab = "Continous x: log(mass of female)",
      ylab = "Continous y: mass of female")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



## Add "smoother"

* "geom = c("point","smooth")"
* Multiple types of smoothers 
    + loess 
    + spline 
    + GAM

```{r, gg.with.smoother, echo = F}
qplot(y = fat.percent,
      x = log(mass.female),
      geom = c("point","smooth"),
      data = milk2,
            main = "Regression data: continous vs log(continuous)",
      xlab = "Continous x: log(mass of female)",
      ylab = "Continous y: mass of female")
```

```{r gg.with.smoother, eval = F}
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



## Look at % milk fat by a categorical variable (diet)

```{r, echo = F}
qplot(y = fat.percent,
      x = diet,
      data = milk2)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



## Look at % milk fat by diet AND biome

* "facets" are a powerful tool in ggplot
* Turns out ALL of our aquatic spp are carnivores (sorry, no manatees)
* This shows that "aquatic" and "diet" are "Confounded"
```{r}
library(ggplot2)

qplot(y = fat.percent,
      x = diet,
      data = milk2,
      facets = ~ biome)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Look at % milk fat by diet AND biome, colored by body mass

* The color scale is DOMINATED by a whale
```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = diet,
      data = milk2,
      color = mass.female,
      size = 2,
      facets = ~ biome)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## ...Look at % milk fat by diet AND biome, colored by log(body mass)

* logging should equalize things

```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = diet,
      data = milk2,
      color = log(mass.female),
      size = 2,
      facets = ~ biome)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Change to boxplot using "geoms"
```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = diet,
      data = milk2,
      geom = c("boxplot"),
      color = log(mass.female),
      facets = ~ biome)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Overlay raw data on boxplots

* geom = c("boxplot", "point")
```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = diet,
      data = milk2,
      geom = c("boxplot", "point"),
      color = log(mass.female),
      facets = ~ biome)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



# Focal data subset: Primates & Relatives

* Primates
* Rodents (this is why we use mouse models)
* Rabbits

<br>

## Subset data

This is a fairly fancy subsetting
* Note use of with()
* which()
* %in%

```{r}
#Select subset
i.use <- with(milk2, which(order %in% c("Rodentia",
                                        "Primates",
                                        "Lagomorpha")))
#make subset using row indexing
milk3 <- milk2[i.use,]

write.csv(milk3, file = "Skibiel_clean_milk_focal_genera.csv")
```

Compare original data and our working subset

* Original = 130 speices
* Working data = 42

```{r}
# Original data
dim(milk2)

#out sowkring subset
dim(milk3)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


# Plot working data: 
## milk fat.percent ~ female size


```{r, echo = F}
library(ggplot2)

qplot(y = fat.percent,
      x = mass.female,
      data = milk3) 
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## The big animal is a Gorrilla!
```{r, echo = F}
library(ggplot2)
milk3$spp.focal <- ""
i.focal <- c(17,18,19,23,30,34,39)

milk3$spp.focal[i.focal] <- as.character(milk3$spp[i.focal])
qplot(y = fat.percent,
      x = mass.female,
      data = milk3) + geom_text(aes(label=spp.focal),hjust=1, vjust=0)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


# Plot logged data
## milk fat.percent ~ log(female size)

```{r, echo = F}
qplot(y = fat.percent,
      x = log(mass.female),
      xlim = c(2,13),
      data = milk3)+ geom_text(aes(label=spp.focal),hjust=0, vjust=0)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Add smoother

% Fat generally declines as body mass increases
```{r, echo = F}
qplot(y = fat.percent,
      x = log(mass.female),
      data = milk3,
      geom = c("point","smooth"))
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Add regression line

* % Fat generally declines as body mass increases
* Grey areas is "confidence band"
* ggplot makes these really easily
* harder using plot()

```{r, echo = F}
qplot(y = fat.percent,
      x = log(mass.female),
      data = milk3,
      geom = c("point")) +
  geom_smooth(method = "lm")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


# Alternate predictor: Duration of gestation
## fat.percent ~ log(gest.month.NUM)

* % Fat generally declines as gestation duration increases
* (the longer animal is pregnant, the less fatty the milk is)
* longer pregnancy, lower milk fat

```{r, echo = F}
qplot(y = fat.percent,
      x = log(gest.month.NUM),
      data = milk3,
      geom = c("point")) +
  geom_smooth(method = "lm")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



# Duration of lactation
## fat.percent ~ log(lacat.mo.NUM)

* % Fat generally declines as lactation duration increases
* longer time kids are dependent, lower milk fat

```{r, echo = F}
qplot(y = fat.percent,
      x = log(lacat.mo.NUM),
      data = milk3,
      geom = c("point")) +
  geom_smooth(method = "lm")
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


# Mass of litter
## fat.percent ~ log(mass.litter)

* % Fat generally declines as litter size  increases
* more or bigger babies, less milk fat

```{r, echo = F}
qplot(y = fat.percent,
      x = log(mass.litter),
      data = milk3,
      geom = c("point")) +
  geom_smooth(method = "lm")
```


