---
title: "One-way ANOVA review"
author: "brouwern@gmail.com"
date: "March 16, 2017"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	eval=FALSE, include=FALSE
)
```

## 1) Load data

* file = simulated_1way_data.csv
* read.csv(file = ...)














```{r, echo = TRUE}
dat<-read.csv("simulated_1way_data.csv")
```

## 2) Plot data in ggplot

### 2a) Basic data visualization

* try different geoms
    + "boxplot"
    + "violin" (violin plot)
    
```{r}
library(ggplot2)
library(cowplot)
```















```{r}
qplot(y = resp.cont,
      x = Treatment,
      shape = Treatment,
      data = dat2) 
```




















```{r 2a.jittering}
#Jitttered
## note: calling the geom within qplot() directly ( qplot(..., geom = "jitter")) doesn't allow control of width for some reason
qplot(y = resp.cont,
      x = Treatment,
      data = dat2) + 
  geom_jitter(width = 0.25)
```




















* trying layering geoms
    + geom = c("boxplot","point")
```{r 2a.layering}
#layering geoms
qplot(y = resp.cont,
      x = Treatment,
      data = dat2,
      geom = c("boxplot",
               "point")) 
```


















### 2b)Plotting means with ggplot

* use ggplot's stat_summary() function to plot means and error bars
    + ggplot(...) + stat_summary()

```{r 2b.stat_summary}
ggplot(data = dat2, 
       aes(y = resp.cont, 
           x = Treatment)) +
stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96))
```



### 2c)Plotting means and rawdata with ggplot

* Use ggplot, stat_summary, and geom_point to plot raw data with means and error bars
    + ggplot(...) + stat_summary() + geom_point()
    + use aes(...) to change colors
    
    
    
    
    
    
    
    
    















```{r}
ggplot(data = dat2, 
       aes(y = resp.cont, 
           x = Treatment)) +
stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96)) +
geom_point(aes(color = Treatment))
```

<br><br><br>

## 3) 1-way ANOVA

### 3a) 1-way ANOVA Global F-test / Omnibus test

* Build a "null model" for these data that representing the null hypothesis using the lm() function
    + ie, "m.null <- lm(resp.cont ~ 1, data = dat)"
* Build a model representing the alternative hypothesis (m.alt)
* Look at these models with the summary() command
    + What do the parameters from m.alt mean?
* Compare these two models with the anova() command
    + What would/does a low p-value for this model comparison mean?
* BONUS: save the output of the anova() command to an object and try to access just the p-value




















```{r}
#null model
m.null <- lm(resp.cont ~ 1, 
             data = dat2)

#alternative model
m.alt  <- lm(resp.cont ~ Treatment, 
             data = dat2)

#model summaries
summary(m.null)
summary(m.alt)

#comare models
## this yields the F statistics
## and the p-value
anova(m.null,
      m.alt)
```











### 3b) Post-hoc comparisons

### Background

* The F-test only tells us that at least 1 mean is different from 1 of the others
* We can interpret out 95% confidence intervals, but it would be good to get p-values for the comparisons
* Because there are 3 means there are 3 possible comparisons
    + Control vs Drug A
    + Control vs. Drug B
    + Drug A vs Drug B
* If we had additional drugs in our experiments the number of possible comparisons woud increase exponentially
* To control for Type I errors (rejecting null that is indeed false) many people (but not all!...) advocate some kind of "correction for multiple comparisons"
* There is an obnoxious multitude of ways to do this.
* There are 2 general categories of approachs:
    + Bonferonni-based methods: pariwise.t.test() in R
    + Parameteric methods: TukeyHSD() in R
    
    
<br><br>    
    
### Bonferonni-based approaches

* Use pairwise.t.test() on your data
    + Note that the syntax is different than lm(); is doesen't use the formula notation with a "~"
    + pairwise.t.test()
    
    
    
    
    
    
    
    
    
    
    
    
    
    

```{r}
pairwise.t.test(x = dat2$resp.cont,
                g = dat2$Treatment,
                pool.sd = T)
```






### Parameteric methods

* Use TukeyHSD()
* TukeyHSD needs a fitted model to work, AND that model must be fit with the aov() function, NOT with lm()
    + aov(resp.cont ~ Treatment, ...)
* Run the test: TukeyHSD(aov.model)
* Plot the output plot(TukeyHSD(aov.model))
















```{r}
# Fit aov() model
m.aov <- aov(resp.cont ~ Treatment, 
         data = dat2)

# Run TukeyHSD
TukeyHSD(m.aov)
```


```{r}
# Fit aov() model
m.aov <- aov(resp.cont ~ Treatment, 
         data = dat2)

# Plot TukeyHSD
plot(TukeyHSD(m.aov))
```

