---
title: "Two-way ANOVA"
author: "brouwern@gmail.com"
date: "March 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#Load data if not already loaded
dat.2way <- read.csv("simulated_2way_data.csv")
```
```{r}
dat.2way$drug.A <- factor(dat.2way$drug.A,
                          levels = levels(dat.2way$drug.A)[c(2,1)])

dat.2way$drug.B <- factor(dat.2way$drug.B,
                          levels = levels(dat.2way$drug.B)[c(2,1)])
```


```{r}
#current factor leevels
levels(dat.2way$Treatment.combo)

#vector re-ordering the factor levels
new.levs <- c("no.A/no.B","drug.A/no.B"
              ,"no.A/drug.B","drug.A/drug.B")

#reset the factor levels
dat.2way$Treatment.combo  <- factor(dat.2way$Treatment.combo,
                                    levels = c(new.levs))
pd <- position_dodge(5)

```


## 1) What are hypotheses?

* There are FIVE hypotheses that can be considered
    + 1 null (Ho) where nothing happens (neither drug has any effect)
    + 4 alternatives with different drug effects
* The null
    + Ho: neither drug has an effect
* 1st set of alternatives: one drug works, the other doesn't
    + Ha1-A: drug A has an effect, but not drug B
    + Ha2-B: drug B has an effect, but not A
* 2nd set of alternatives: both drugs have an effect
    + Ha3-A+B: both drugs have effects, but they are sperate and independent (additive model)
    + Ha4-A*B: both drugs have an effect, and the presence of one changes the effect of the other (multiplicative model)
    

## 2) Set up models in R

```{r}
#1)Null model: 
#"Intercept only" (grand mean onnly)
m.null         <-lm(resp.cont ~ 1, data = dat.2way)

#2)drug.A only model: 
# Intercept + treatment effect
#Typical R code
m.A.only      <- lm(resp.cont ~ drug.A,  data = dat.2way)

#Equivalent: 
## Include the "1 + ..."" for the intercept
## Intercept is normally implied
m.A.only      <- lm(resp.cont ~ 1+ drug.A,  data = dat.2way)


#3)drug.A only model: 
#Intercept + treatment effect
m.B.only      <- lm(resp.cont ~          drug.B, data = dat.2way)


#4)drug.A and drug.B model: 
#"Additive model": independent of effects of each drug
m.ADDITIVE    <- lm(resp.cont ~ drug.A + drug.B, data = dat.2way)


#5)drug.A*drug.B model
#"interaction model": one drug changes the effect of the other

#Typical R formulation
m.interaction <- lm(resp.cont ~ drug.A * drug.B, data = dat.2way)

#Equivalent:
### Additive terms normally implied
m.interaction <- lm(resp.cont ~ drug.A + drug.B + drug.A * drug.B, data = dat.2way)


```



## Single model

```{r}
anova(m.interaction)
```


### 3) Make a plot representing the null model

```{r}
summary(m.null)
```


```{r}
dat.2way$null.model <- 1
ggplot(aes(y=resp.cont,
      x = null.model),
      data = dat.2way) +
 stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96),
             size = 2) +
  geom_point(aes(color = Treatment.combo))
```



## 4) Make a plot for Drug-A only model

```{r}
summary(m.A.only)
```


```{r}
ggplot(aes(y=resp.cont,
      x = drug.A),
      data = dat.2way) +
 stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96),
             size = 2) +
  geom_point(aes(color = Treatment.combo),
             size = 4)
```






# B-only

```{r}
summary(m.B.only)
```

```{r}
ggplot(aes(y=resp.cont,
      x = drug.B),
      data = dat.2way) +
 stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96),
             size = 2) +
  geom_point(aes(color = Treatment.combo),
             size = 4)
```


### Additive model

```{r}
summary(m.ADDITIVE)
```

```{r}
dat.2way$Treatments.additive <- as.character(dat.2way$Treatment.combo)
dat.2way$Treatments.additive <- ifelse(dat.2way$Treatments.additive == "no.A/no.B",
                                       "neither.drug",
                                       dat.2way$Treatments.additive)
dat.2way$Treatments.additive <- ifelse(dat.2way$Treatments.additive == "drug.A/no.B",
                                       "drug.A.only",
                                       dat.2way$Treatments.additive)
dat.2way$Treatments.additive <- ifelse(dat.2way$Treatments.additive == "no.A/drug.B",
                                       "drug.B.only",
                                       dat.2way$Treatments.additive)

dat.2way$Treatments.additive <- ifelse(dat.2way$Treatments.additive == "drug.A/drug.B",
                                       "both.drugs",
                                       dat.2way$Treatments.additive)

dat.2way$Treatments.additive <- factor(dat.2way$Treatments.additive)

summary(dat.2way$Treatments.additive)

dat.2way$Treatments.additive <- factor(dat.2way$Treatments.additive,
                                       levels = levels(dat.2way$Treatments.additive)[c(4,2,3,1)])

summary(dat.2way$Treatments.additive)
```

```{r}


ggplot(aes(y=resp.cont,
      x = Treatments.additive),
      data = dat.2way) +
 stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96),
             size = 2) +
  geom_point(aes(color = Treatment.combo),
             size = 4)
```


```{r}
newdat <- expand.grid(drug.A = levels(dat.2way$drug.A),
                      drug.B = levels(dat.2way$drug.B))

pred.out <- predict(m.ADDITIVE,
                    newdata = newdat,
                    se.fit = F,
                    interval = "confidence")

newdat <- cbind(newdat, pred.out)

newdat$treatment.combo <- c("neither.drug",
                            "drug.A.only",
                            "drug.B.only",
                            "both.drugs")

newdat$treatment.combo <- factor(newdat$treatment.combo)
newdat$treatment.combo <- factor(newdat$treatment.combo,
                           levels  = levels(newdat$treatment.combo)[c(4,2,3,1)])

dat.2way$treatment.combo <- dat.2way$Treatments.additive
qplot(y = fit,
      x = treatment.combo,
      data = newdat) +
  geom_point(size = 4)+
  geom_errorbar(aes(ymin = lwr,
                    ymax = upr),
                width = 0) +
  geom_point(data = dat.2way,
             aes(y = resp.cont,
                 x = treatment.combo,
                 color = treatment.combo))



```

```{r}
summary(m.ADDITIVE)
```


```{r}
summary(m.interaction)
```

```{r}
newdat <- expand.grid(drug.A = levels(dat.2way$drug.A),
                      drug.B = levels(dat.2way$drug.B))

pred.out <- predict(m.interaction,
                    newdata = newdat,
                    se.fit = F,
                    interval = "confidence")

newdat <- cbind(newdat, pred.out)

newdat$treatment.combo <- c("neither.drug",
                            "drug.A.only",
                            "drug.B.only",
                            "both.drugs")

newdat$treatment.combo <- factor(newdat$treatment.combo)
newdat$treatment.combo <- factor(newdat$treatment.combo,
                           levels  = levels(newdat$treatment.combo)[c(4,2,3,1)])

qplot(y = fit,
      x = treatment.combo,
      data = newdat) +
  geom_point(size = 4)+
  geom_errorbar(aes(ymin = lwr,
                    ymax = upr),
                width = 0) +
    geom_point(data = dat.2way,
             aes(y = resp.cont,
                 x = treatment.combo,
                 color = treatment.combo))

```




```{r}
library(phia)

interactionMeans(m.interaction)

```



m.null
m.A.only
m.B.only
m.ADDITIVE
m.interaction


```{r}
anova(m.null, m.B.only)
```


```{r}
anova(m.null, m.A.only)
```

```{r}
anova(m.ADDITIVE, m.A.only)
```


```{r}
anova(m.ADDITIVE, m.B.only)
```


```{r}
anova(m.interaction, m.ADDITIVE)
```



```{r}
anova(m.interaction)
```


```{r}
library(MASS)
Anova(m.interaction)
```



```{r}
library(MASS)
Anova(m.interaction, type = "III")
```

