---
title: "Untitled"
author: "brouwern@gmail.com"
date: "April 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 
# Vocab

* exponential family of distributions
    + normal/gaussian: 
    + binomial: 0/1, yes/no, present/absent, successes/trials
    + poisson: counts
    + negative binomial: counts (neither negative, nor binomial)
    + gamma: data w/positive skew
    + beta (?): not exponential
* link function
* inverse link function
* cannonical link function
* odds ratio
* log odds ratio
* rate ratio
* back transformation of coefficients




# Load data

```{r}
marbles <- read.csv("Lazic_weight_marbles.csv")
```



```{r}
library(ggplot2)
ggplot(aes(y = marbles,
      x = weight),
      data = marbles) +
  geom_smooth(method = glm,se = FALSE,
              method.args = list(family = "poisson"),
              size = 2) +
  geom_point(aes(color = sex, shape = sex),size =3)
```

# Distributions


## Normal

code for curve:
http://stackoverflow.com/questions/6967664/ggplot2-histogram-with-normal-curve

```{r}
par(mfrow = c(1,1))
marbles <- read.csv("Lazic_weight_marbles.csv")
marbles.noNA <- na.omit(marbles)
i.females <- which(marbles.noNA$sex == "Female")
marbles.noNA.f <- marbles.noNA[i.females,]
library(ggplot2);library(cowplot);library(plyr)





mean(marbles.noNA.f$weight)
sd(marbles.noNA.f$weight)

```


```{r}
binwidth <- 1
qplot(weight,
      data =marbles.noNA.f) + 
  geom_vline(xintercept = mean(marbles.noNA.f$weight),
             color = "red", 
             size = 3) +
  xlim(0,25)+
  stat_function( 
    fun = function(x, mean, sd, n, bw){ 
      dnorm(x = x, mean = mean, sd = sd) * n * bw
    }, 
    args = c(mean = mean(marbles.noNA.f$weight), 
             sd = sd(marbles.noNA.f$weight), 
             n = dim(marbles.noNA.f)[1], 
             bw = binwidth))
```




## Poisson


Raw data
```{r}
mean(marbles.noNA.f$marbles)
table(marbles.noNA.f$marbles)
sd(marbles.noNA.f$marbles)
var(marbles.noNA.f$marbles)

qplot(marbles,
      data =marbles.noNA.f) + 
  geom_vline(xintercept = mean(marbles.noNA.f$marbles),
             color = "red", 
             size = 3) 
```


## count data w/normal density

```{r}
binwidth <- 2
qplot(marbles,
      data =marbles.noNA.f) + 
  geom_vline(xintercept = mean(marbles.noNA.f$marbles),
             color = "red", 
             size = 3) +
  stat_function( 
    fun = function(x, mean, sd, n, bw){ 
      dnorm(x = x, mean = mean, sd = sd) * n * bw
    }, 
    args = c(mean = mean(marbles.noNA.f$marbles), 
             sd = sd(marbles.noNA.f$marbles), 
             n = dim(marbles.noNA.f)[1], 
             bw = binwidth),
    size = 2,
    color = "blue") +
  xlim(-5,10)
```



## count data w/ poisson
```{r}
binwidth <- 2
x <- 0:10
y <- dpois(seq(0,10), lambda = mean(marbles.noNA.f$marbles))
df<-data.frame(x,y)
qplot(marbles,
      data =marbles.noNA.f) + 
  geom_vline(xintercept = mean(marbles.noNA.f$marbles),
             color = "red", 
             size = 3) +
geom_line(aes(y = y*30,x =x),data = df,
          color = "blue",size = 3)
```




## count data w/ negative binomial
```{r}
binwidth <- 2
x <- 0:10
y <- dnbinom(seq(0,10), 
             mu = 1.2321,
             size = 1.244)
df<-data.frame(x,y)
qplot(marbles,
      data =marbles.noNA.f) + 
  geom_vline(xintercept = mean(marbles.noNA.f$marbles),
             color = "red", 
             size = 3) +
geom_line(aes(y = y*15,x =x),data = df,
          color = "blue",size = 3)
```





# A poisson model

```{r}

glm.pois.mean <- glm(marbles ~ weight, 
          data = marbles,
          family = poisson(link = "log"))



summary(glm.pois.mean)

```

# Overdispersion test

```{r}

deviance(glm.pois.mean)/glm.pois.mean$df.residual



library(AES)
dispersiontest(glm.pois.mean)


```


# Accomodating overdispersion

## Quasi poisson

```{r}

glm.quasipois.mean <- glm(marbles ~ weight, 
          data = marbles,
          family = quasipoisson)



summary(glm.quasipois.mean)
```



## negbin

```{r}
library(MASS)
glm.neginb.mean <- glm.nb(marbles ~ weight, 
          data = marbles)


summary(glm.neginb.mean)

```




# Individual random effct
```{r}

library(lme4)
marbles$row.ID <- 1:dim(marbles)[1]
glmm.indranef <- glmer(marbles ~ weight + (1|row.ID), 
                      family = poisson,
                      data = marbles)


summary(glmm.indranef)



```











## Re-code for logistic regression

```{r}
marbles$marbles01 <- ifelse(marbles$marbles > 0,1,0)

max(marbles$marbles)


glm.binom <- glm(cbind(marbles, 16) ~ weight, 
                                      data = marbles, 
                                      family = binomial)


summary(glm.binom)

```





## Binomial glm


```{r}
glm()
```


# Logistic regression - categoricalv ariables


```{r}
mat <- as.matrix(table(marbles$marbles01,
      marbles$sex))

mat
chisq.test(mat,correct = F)
```



```{r}


logit.sex <- glm(marbles01 ~ sex, 
               family = binomial,
               data = marbles)

anova(logit.sex,
      test = "Rao")
```


# Logistic regression - continous variables
```{r}
#remove NAs
marbles.noNA <- na.omit(marbles)

#Run model
logit.weight <- glm(marbles01 ~ weight, 
               family = binomial,
               data = marbles.noNA)


#test model
anova(logit.weight,
      test="Chisq")

anova(logit.weight,
      test="LRT")

confint(logit.weight)


```




# predictions

```{r}

marbles.noNA$prediction.logit <- predict.glm(logit.weight)


SE.out <- predict.glm(logit.weight,
                      se.fit = TRUE)

marbles.noNA$prediction.SE <- SE.out$se.fit





marbles.noNA$prediction.prob <- predict.glm(logit.weight,
                                             type = "response")

marbles.noNA$SE.lo <- invlogit(marbles.noNA$prediction.logit -
                               1.96*marbles.noNA$prediction.SE )

marbles.noNA$SE.up <- invlogit(marbles.noNA$prediction.logit +
                               1.96*marbles.noNA$prediction.SE )







```



```{r}
ggplot(aes(y =  prediction.prob,
          x = weight),
      data = marbles.noNA) +
  geom_line(size = 4, color = "blue")  +
  geom_line(aes(y=SE.up,
                x = weight)) +
  geom_line(aes(y=SE.lo,
                x = weight)) +
  geom_hline(yintercept = 1,linetype = 2)
```




# Look for overdispersion
```{r}
quasilogit.weight <- glm(marbles01 ~ weight, 
               family = quasibinomial,
               data = marbles.noNA)

```



```{r}
summary(logit.m1)
```






# Easy logistic regression plot

## One continous predictor variable
```{r}
library(ggplot2)

qplot(y = marbles01,
      x = weight,
      data = marbles) +
  geom_point(size = 3) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = FALSE) +
  xlab("Continous covariate: weight") +
  ylab("Binary response: 0 or 1") +
  ggtitle("Logistic regression with continuous covariate")

```


## One continous and one categorical variable

Discrete groups can be split up by defining a color
```{r}
library(ggplot2)

qplot(y = marbles01,
      x = weight,
      color = drug,
      linetype = drug,
      data = marbles) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = FALSE) +
  xlab("Continous covariate: weight") +
  ylab("Binary response: 0 or 1") +
  ggtitle("Logistic regression plot with ggplot2")

```




## One continous and multiple categories

Combining categorial variables allows you to plot further interactions.


```{r}
library(ggplot2)

marbles$group.x.drug <- with(marbles,
                           paste(group,
                                 drug,
                                 sep = "."))

qplot(y = marbles01,
      x = weight,
      color = group.x.drug,
      linetype = group.x.drug,
      data = marbles,
      facets = group ~ drug) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = FALSE) +
  xlab("Continous covariate: weight") +
  ylab("Binary response: 0 or 1") +
  ggtitle("Logistic regression plot with ggplot2") +
  

```












