---
title: "Lazic marble burying assay"
author: "brouwern@gmail.com"
date: "April 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Load data

Load Lazic marble burying data
```{r}
marbles <- read.csv("Lazic_weight_marbles.csv")
```

# Recode as 0/1 for logistic regression

```{r}
marbles$marbles01 <- ifelse(marbles$marbles > 0,1,0)
```


# Plot poisson regression line

marbles buried vs. weight of rat

```{r}
library(ggplot2)
ggplot(aes(y = marbles,
      x = weight),
      data = marbles) +
  geom_smooth(method = glm,
              se = FALSE,
              method.args = list(family = "poisson"),
              size = 2) +
  geom_point(aes(color = sex, 
                 shape = sex),
             size =3)
```


# Model: Poisson regression with continous predictor, weight

```{r}
glm.pois.mean <- glm(marbles ~ weight, 
          data = marbles,
          family = poisson(link = "log"))

```


# Logistic regression with all 4 treatment combos

```{r}
glm.logit.full <- glm(marbles01 ~ group*drug,
                      family = binomial,
                      data = marbles)
```

# Plot predictions

Make fake data
```{r}
summary(marbles)
fake.data <- expand.grid(group = c("SAL","VPA"),
                         drug = c("MPEP","SAL"))

fake.data
```

Put fake data into predictions
```{r}
fake.data$logit.preds <- predict(glm.logit.full,
                                 newdata = fake.data)

```


SE of predictions
```{r}
SE.preds <- predict(glm.logit.full,
                                 newdata = fake.data,
                                 se.fit = TRUE)

fake.data$SE <- SE.preds$se.fit
```


Predictions on probablity scale
```{r}
fake.data$real.pred <- invlogit(fake.data$predictions)
```


95% CI on probalbity scale
```{r}
fake.data$CI.lo <- invlogit(fake.data$predictions -
                            1.96*fake.data$SE)

fake.data$CI.hi <- invlogit(fake.data$predictions +
                            1.96*fake.data$SE)
```

Plot output

```{r}
fake.data$treatment <- paste(fake.data$group,
                             "\n",
                             fake.data$drug,
                             sep = "")
pd <- position_dodge(0.2)
qplot(y = real.pred,
      x = treatment,
      color = drug,
      shape = group,
      data = fake.data) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI.lo,
                    ymax = CI.hi),
                width = 0)
```


