---
title: "Intro To ANOVA 1: Make random data"
author: "brouwern@gmail.com"
date: "March 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
fw <- 1
```

# An Introduction to ANOVA

* Krywinski & Altman (2014) present an analysis simulated data 
* We'll use their parameters to carry out a similar analysis

Reference: Krywinski & Altman (2014).  Designing comparative experiments.  Nature Methods 11:597-8.


# Outline

* Define parameters for data simulation
* Generate random data
* Organize data into dataframe





# The hypothetical study

* Experimental study of cells fluorescing after treatment
* Control (no treatment)
* Treatment A
* Treatment B
* each treatment has n = 17 indepndent samples.  


<br>


# Basic parameters for simulation:

Data are defined by the following parameters

* control = Background rate of fluorescence of cells
    + = "mu", the Greek letter for "m"; looks like a "u"
* change in fluoresence due to treatments A and B
* Variability in fluorescence (random variation, "noise"")







## Means

### Mean of control

The overall baseline or background rate of fluroescence

```{r mu}
mu <- 10   #control
```


### Treatment effects

How much the treatments ("trt") cause the samples to diverge or differ from means

```{r effects}
#d is for "difference"
dA <- 0.6  #trt A effect
dB <- 1    #trt B effect
```

### Treatment means

* ie Mean of treatment A 
    + = (background) + (treatment B effect)
    + = (control)    + dB
    + = mu + dB
    
```{r all.means}
muC <- mu      #control
muA <- mu + dA #trt A
muB <- mu + dB #trt B
```


## Variablity around mean

Variation around mean is assumed to be normally distributed with a constant variance

```{r}
s  <- 1 #standard deviation
```


## Sample Size

```{r}
n <- 17
```



# Generating random data

* rnorm() function in R
* 3 arguments:
    + n = number of replications 
    + mean
    + sd = standard deviation


## Generate random data

```{r rnorm}
#generate random control data
dataC <- rnorm(n = n, 
               mean = muC,
               sd = s)

```


## Visualize random data

We'll use ggplot

* ggplot2:qplot
    + geom = "histogram"
    + no "y = "
    + won't work: "qplot(y = dataC, geom = "histogram")"
    
```{r, message=F, warning=F, fig.width = 3}
library(ggplot2)
library(cowplot)

qplot(dataC,
      geom = "histogram")
```




## Random data for other treatments
```{r}
#data for treatment A
dataA <- rnorm(n = n, 
               mean = muA,
               sd = s)


#data for treatment B
dataB <- rnorm(n = n, 
               mean = muB,
               sd = s)
```






# Dataframe

* Put data into a dataframe using data.frame()
* Must be in "stacked" format
    + All y variables in a single column
    + 2nd column with indicator of the treatment
* I will use the rep() function to help with this

## Put all y variables into single vector
```{r}
y.column <- c(dataA,
              dataC, 
              dataB)

#check length
length(y.column)

```


## Make the 1st treatment vector

Reapt the letter "A" n = 17 times

```{r, include=FALSE}
#letter designating the treatment
vectorA <- rep("A",n)

#This is same as this
## (Don't have to to this!)
vectorA <- c("A","A","A","A","A",
             "A","A","A","A","A",
             "A","A","A","A","A",
             "A","A")


#Check the length
length(vectorA)
```



## The the other vectors
```{r}
#Make other vectors
## rep C for control
vectorC <- rep("C",n)

## rep B for treatment B
vectorB <- rep("B",n)


```


Put all of the treatment vectors into 1 vector that will become the "treatment" column of our dataframe

```{r}
trt.column <- c(vectorA,
                vectorC,
                vectorB)
```



## Make dataframe

Dataframe with 2 columns

* fluoro = y variable (respones variable)
* treatment = indicator of treatment group (A,B, C)

```{r}
dat <- data.frame(fluoro = y.column,
                  treatment = trt.column)
```


To save space when we print results, round the fluoro variable
```{r}
dat$fluoro <- round(dat$fluoro, 
                    digits = 2)
```


Look at dataframe

```{r}
head(dat)
```


Summary of dataframe

```{r}
summary(dat)
```


## Summary statistics by group

* doBy package
* summarBy() function
* FUN = what function to use
* FUN = c(mean,sd) applies both functions at the sample time

```{r}
library(doBy)


summaryBy(fluoro ~ treatment, 
          data = dat, 
          FUN = c(mean,sd))
```

