---
title: "Intro To ANOVA 2: Explore data with ggplot"
author: "brouwern@gmail.com"
date: "March 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T)
```

# Outline

* Plot data in boxplot wiht ggplot2:qplot
* Explore functionality of qplot
    + color, shape
    + "jittering"
* Change factor levels using factor()
    + change "factor levels"
    + change factor names/labels
* Plot means

# Make data that matches Nature Methods Paper

```{r}
y.column <- c(	11.24329808,	10.57397347,	8.798111146,	9.648373069,	10.16202242
,	9.253887088,	11.51807329,	12.22480837,	8.797802989,	10.6733833,	12.38641145,	10.55536801,	10.83550408,	9.473762458,	11.8490472,	11.52501837,	10.02532039,	10.26283115,	9.541543832,	10.8289181,	10.61879171,	10.56255483,	10.12447501,	11.27377454,	8.985686839,	9.661441062,	9.337572633,	9.624627193,	9.562961301,	8.898324866,	9.033531311,	12.00679606,	10.80340919,	10.72436056,	10.39465146,	10.29051717,	11.32943119,	11.95522121
,	12.59729367,	10.49891019,	9.58186603,	10.83069437,	9.639272696,	11.23721842,	10.25767478,	10.83731922,	10.46711202,	10.6520578,	11.90867276,	12.38906162,	11.33769377)

n <- 17
#create stacked dataframe
dat <- data.frame(fluoro = y.column,
                  treatment = c(rep("A",n),rep("C",n),rep("B",n)))

#set factor levels
dat$treatment <- factor(dat$treatment ,
                        levels = c("A","C","B"))
```

# Histograms

```{r}
library(ggplot2)
qplot(dat$fluoro)
```




Make panel a) in published figur


```{r, fig.width = 4}
library(ggplot2)
qplot(fluoro,
      geom = "density",
      color = treatment,
      data = dat)
```


CHange x axis limits

```{r, fig.width =5}
library(ggplot2)
qplot(fluoro,
      geom = "density",
      color = treatment,
      data = dat,
      xlim = c(5,14.5))
```



# Create "facets"

```{r, fig.width=4}
library(ggplot2)
qplot(fluoro,
      geom = "density",
      color = treatment,
      data = dat,
      xlim = c(7,14.5),
      facets = . ~ treatment)
```


```{r, fig.width=5}
library(ggplot2)
qplot(fluoro,
      geom = "density",
      color = treatment,
      data = dat,
      xlim = c(7,14.5),
      facets = treatment ~ .)
```



Add fill


```{r}
library(ggplot2)
qplot(fluoro,
      geom = "density",
      color = treatment,
      data = dat,
      xlim = c(7,14.5),
      fill = treatment,
      facets = treatment ~ .)
```




# Boxplots

## Raw data
```{r, fig.width = 5}
library(ggplot2)

qplot(y = fluoro,
      x = treatment,
      data = dat)
```


## Boxplot

* geom = "boxplot"

```{r, fig.width = 5}
library(ggplot2)

qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = "violin")
```


## Boxplot w/raw data overlay

* Can include 2 geoms
* geom = c("boxplot","point")
    + note: must be "point", NOT "points" w/an "s"

```{r, fig.width = 5}
library(ggplot2)

qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot",
               "point"))
```



## Change colors

* color = treatment
```{r, fig.width = 5}
qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot",
               "point"),
      color = treatment)
```


## Change point shape

* shape = treatment
```{r, fig.width = 5}
qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot",
               "point"),
      color = treatment,
      shape = treatment)
```



## "jitter" points

* geom = c(..., "jitter")

```{r, fig.width = 5}
qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot",
               "jitter"),
      color = treatment,
      shape = treatment)
```



## External geoms

* remove geom = c(..., "jitter")
* add geom_jitter() outside of qplot
* note use of "+" sign

```{r, fig.width = 5}
qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot"), #no jitter
      color = treatment,
      shape = treatment) + #note "+"
  geom_jitter() # <=jitter added here
```

## Chaning amont of jitter

* geom_jitter() takes the argument "width"

```{r, fig.width = 5}
qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = c("boxplot"), #no jitter
      color = treatment,
      shape = treatment) + #note "+"
  geom_jitter(width = 0.1) #jitter added here
```




# Setting factor levels

* To faciliate comparisons, put our control "C" in the middle
* the "treatment" columns is called a "categorical" variable
* aka "factor" variable
* A, B, C, are the "levels" or "factor levels"
* R defaults to alphabetical order
* we can "relevel" them


Note the order when we use summary()
```{r}
summary(dat)
```


## relevel treatment factor

* factor() takes two argument
    + the column to convert
    + the order that the EXISTING factor levels should be changed to

```{r}
dat$treatment <- factor(dat$treatment ,
                        levels = c("A",
                                   "C",
                                   "B"))
```


Now look at summary
```{r}
summary(dat)
```



## Renaming factor levels

* factor also has the argument "labels"
* can change what each level is caleld

```{r}
dat$treatment <- factor(dat$treatment ,
                        labels = c("treatment.A",
                                   "treatment.C",
                                   "treatment.B"))
```


use summary
```{r}
summary(dat)
```


Check out impact on plots
```{r, fig.width = 5}
library(ggplot2)

qplot(y = fluoro,
      x = treatment,
      data = dat,
      geom = "boxplot")
```



# Plots of means

Plot means w/error bars in ggplot

* We could do this by first summarizing the data 
    + calculate mean
    + calcualte sd
    + calcualte se from sd
    + make a table
    + plot it
* ggplot can do this all "on the fly"
    + stat_summary
    
    

# First - a function for the standard error

* basic formula: se = sd/sqrt(n)
* R code:        se = sd(x)/sqrt(n)

```{r}
se <- function(x){
  n <- length(x)
  sd(x)/sqrt(n)
}
```


# The ggplot command

Very powerful, a bit complex....

* Key part: ggplot(data = ..., aes(....) )
* the stuff that goes in aes(...) is really really important!
* aes = "aesthetic"

```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment))

```

## What happened?

* qplot, you declare geoms inside the call to qplot
    + qplot(, geom = c(...))
    + default is geom = "point"
* ggplot
    + no default
    + geoms go OUTSIDE of main ggplot call
    + each geom has its own fuction
    + ie, geom_point, geom_histogram, geom_boxplot
    
    

# ggplot(...) + geom_point

* note "+" outside of main ggplot() call

```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + #"+" 
  geom_point()

```



## ggplot + geom_boxplot()


```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + #"+" 
  geom_boxplot()

```

## ggplot + geom_boxplot() + geom_point()


```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + #"+" 
  geom_boxplot() +
  geom_point()

```


# Changing aesthetic


## Changing global aesthetic

* add "color = treatment" within the main ggplot aes()
* note that BOTH boxplots AND point changed
* with have set the aesthetic for ALL Of the "layers"

```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment,
           color = treatment)) + 
  geom_boxplot() +
  geom_point()

```

## Changing individual layer aesthetic

### Change just boxplotes
* add "color = treatment" w/in geom_boxplot()
* boxes colored, points black!

```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + 
  geom_boxplot(color = c("red","green","blue")) +
  geom_point()

```



### Change just points
* add "color = treatment" w/in geom_point()
* boxes colored, points black!

```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + 
  geom_boxplot() +
  geom_point(aes(color = treatment,
                 shape = treatment))

```




# geom stat_summary()

* the deafults for stat_summary() are
    + plot mean as a point
    + plot error bars as +/- 1 se
```{r, fig.width = 5}
library(ggplot2)
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + 
  stat_summary() 

```




# To fully write out what the default

* the stat_summary() function has an arguement
    + fun.data = "mean_se"
    

```{r,fig.width = 5}
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + 
stat_summary(fun.data = "mean_se")
```



## Plotting 95% CIs

* fun.data is a function with an argument "mult"
* deafult is 1
* so ggplot plots +/- 1*SE
* what happens when we change this to 1.96?
* requires a little bit of code...
    + fun.args = list(mult = 1.96)
    
```{r,fig.width = 5}
ggplot(data = dat, 
       aes(y = fluoro, 
           x = treatment)) + 
stat_summary(fun.data = "mean_se",
             fun.args = list(mult = 1.96))
```





# Make a function to make this easy
```{r}
easy.ggplot.ci <- function(dat, y, x, mult = 1){
  ggplot(data = dat, 
       aes(y = dat[,y], 
           x = dat[,x])) + 
stat_summary(fun.data = "mean_se",
             fun.args = list(mult = mult)) +
    xlab(x)+ylab(y)
}
```




# Plot error bars as SE

* the deafult is mult = 1
* this gives us +/- 1*SE
```{r,fig.width = 5}
easy.ggplot.ci(dat = dat, 
         y = "fluoro",
         x = "treatment")
```

# Plot error bars as 95% CI

* change mult = 1.96
```{r,fig.width = 5}
easy.ggplot.ci(dat = dat, 
         y = "fluoro",
         x = "treatment",
         mult = 1.96) 
```
