---
title: "Intro To ANOVA 3: t-tests"
author: "brouwern@gmail.com"
date: "March 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Make data that matches Nature Methods Paper

```{r}
y.column <- c(	11.24329808,	10.57397347,	8.798111146,	9.648373069,	10.16202242
,	9.253887088,	11.51807329,	12.22480837,	8.797802989,	10.6733833,	12.38641145,	10.55536801,	10.83550408,	9.473762458,	11.8490472,	11.52501837,	10.02532039,	10.26283115,	9.541543832,	10.8289181,	10.61879171,	10.56255483,	10.12447501,	11.27377454,	8.985686839,	9.661441062,	9.337572633,	9.624627193,	9.562961301,	8.898324866,	9.033531311,	12.00679606,	10.80340919,	10.72436056,	10.39465146,	10.29051717,	11.32943119,	11.95522121
,	12.59729367,	10.49891019,	9.58186603,	10.83069437,	9.639272696,	11.23721842,	10.25767478,	10.83731922,	10.46711202,	10.6520578,	11.90867276,	12.38906162,	11.33769377)


#create stacked dataframe
n<-17
dat <- data.frame(fluoro = y.column,
                  treatment = c(rep("A",n),rep("C",n),rep("B",n)))

#set factor levels
dat$treatment <- factor(dat$treatment ,
                        levels = c("A","C","B"))
```


# Outline

* Create indicies that define subsets of the data
    + using %in%
* Do t-tests on subsets of data


## Creating subsets

* This can be done several ways
* I will do it using indexing
* subset() command also works well


### Isolate "A" and "C" treatment


#### Slow way: using which() twice

3 steps: 

* locate As indices
* locate Cs 
* combine indices

```{r}
#Which rows are treatment.A in?
i.A <- which(dat$treatment == "treatment.A")

#Look at the index
i.A
```

#### Use the index using "row indexing"

* Put index into square brackets [ , ]
* [rows, columns]
* [i.A, ] = "give me the rows that match i.A", and all columns


```{r}
dat[i.A, ]
```


We can do this and get just the summary info on A
```{r}
summary(dat[i.A, ])
```



What if I want just the "fluoro" column?

```{r}
dat[i.A, "fluoro"]
```


This can be confusing, but the previous code is equivalent to

```{r, eval=FALSE, include=FALSE}
dat$fluor[i.A]
```


### Now get the controls

How do we isolate the indices for "treatment.C"?

```{r}





















i.C <- which(dat$treatment == "treatment.C")
```


### Put As and Cs together

How do we put the i.A and i.C vector together into a single vector of indices (call it i.AC)?

```{r}


















i.AC <- c(i.A, i.C)
```

### Summary info on A and C
```{r}
summary(dat[i.AC, ])
```


# Fast way to get these indices: %in%

* "which items in $treatment" match the 2 valeus %in% the vector c("A","C") "

```{r}
i.AC <- which(dat$treatment %in% c("A","C"))
```



Call summary() using this new index i.AC

```{r}
















summary(dat[i.AC, ])
```





# Make indices for other items

* i.AB
* i.BC

```{r}


























i.AB <- which(dat$treatment %in% c("A","B"))

i.BC <- which(dat$treatment %in% c("B","C"))
```









# T-tests

## Treatment A vs. treatment C

Run the t-test
```{r}
t.test.AvsC <- t.test(fluoro ~ treatment, 
       dat = dat,
       subset = i.AC)
```


Call up the output

```{r}
t.test.AvsC
```


## T test on the other combinations
```{r}
t.test.AvsB <- t.test(fluoro ~ treatment, 
       dat = dat,
       subset = i.AB)




t.test.BvsC <- t.test(fluoro ~ treatment, 
       dat = dat,
       subset = i.BC)


```


## p values for each one
```{r}
pAvsB <- t.test.AvsB$p.value
pAvsC <- t.test.AvsC$p.value
pBvsC <- t.test.BvsC$p.value

pAvsB
pAvsC
pBvsC
```

## Make a table to hold the p values

Put all of the p values into a vector called "p.valeus"
```{r}
# p.values <- 



















p.values <- c(pAvsB,
              pAvsC,
              pBvsC)
```




Make a vector that holds the names of the comparisons
```{r}
#comparison <- 

















comparison <- c("AvsB","AvsC","BvsC")
```


Make a dataframe
```{r}
p.table <- data.frame(comparison = comparison,
                      p = p.values)

p.table
```


Round the p values

```{r}
#p.table$p <-














#do the rounding
p.table$p <- round(p.table$p,3)

#look at the table
p.table
```



