---
title: "Intro To ANOVA 4: 1-way ANOVA"
author: "brouwern@gmail.com"
date: "March 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Data
```{r}
y.column <- c(	11.24329808,	10.57397347,	8.798111146,	9.648373069,	10.16202242
,	9.253887088,	11.51807329,	12.22480837,	8.797802989,	10.6733833,	12.38641145,	10.55536801,	10.83550408,	9.473762458,	11.8490472,	11.52501837,	10.02532039,	10.26283115,	9.541543832,	10.8289181,	10.61879171,	10.56255483,	10.12447501,	11.27377454,	8.985686839,	9.661441062,	9.337572633,	9.624627193,	9.562961301,	8.898324866,	9.033531311,	12.00679606,	10.80340919,	10.72436056,	10.39465146,	10.29051717,	11.32943119,	11.95522121
,	12.59729367,	10.49891019,	9.58186603,	10.83069437,	9.639272696,	11.23721842,	10.25767478,	10.83731922,	10.46711202,	10.6520578,	11.90867276,	12.38906162,	11.33769377)


#create stacked dataframe
n<-17
dat <- data.frame(fluoro = y.column,
                  treatment = c(rep("A",n),rep("C",n),rep("B",n)))

#set factor levels
dat$treatment <- factor(dat$treatment ,
                        levels = c("A","C","B"))
```


# Plotting Function

```{r}
easy.ggplot.ci <- function(dat, y, x, mult = 1){
  ggplot(data = dat, 
       aes(y = dat[,y], 
           x = dat[,x])) + 
stat_summary(fun.data = "mean_se",
             fun.args = list(mult = mult)) +
    xlab(x)+ylab(y)
}
```


# 1-way ANOVA in R


## Fit null model

* ANOVA, t-test, regression are all linear models
* ~1 means "fit the mean"

```{r}
model.null <- lm(fluoro ~ 1, 
             data = dat)
```



Look at output
```{r}
summary(model.null)
```


The (Intercept) is the same as the grand mean of the data.  Check this with summary.

```{r}
summary(dat)
```




## Fit alt model linear model


```{r}
model1 <- lm(fluoro ~ treatment, 
             data = dat)
```


## Look at summary

* to save space I will use an alternative command "display"
* from arm package

```{r}
library(arm)
display(model1, detail = TRUE)
```


## Compare summary to plot

* treatment A serves as baseline
* treatment C is lower than A
    + so treatmenttreatment.C  is negative (-)
* treatment B is higher than A
    + so treatmenttreatment.B is positive (+)
* p values for treatmenttreatment.C and treatmenttreatment.B should match t-tests
    + check out p.table object


```{r,fig.width = 5}
library(ggplot2)
easy.ggplot.ci(dat = dat, 
         y = "fluoro",
         x = "treatment",
         mult = 1.96)
```


# "Re-parameterize"" model

* This is sometiems useful, but it NOT used for hypothesis testing!
* ... ~ -1 + ... "Drops" the intercept
* we now get the means of each group
* QUESTION: how do we interpret the small p-values?

```{r}
model.means <- lm(fluoro ~ -1 + treatment, 
             data = dat)

display(model.means, detail = T)
```



# ANOVA by model comparison

* the anova() command compares 1 models

Run a model comparison on model.null and model1
```{r}























anova(model.null, 
      model1)
```

This is how I like to think of things: in terms of comparing the fit of 2 models



# ANOVA by "partitioning the variance"

* You can do this too, and will see some people do this.
* Almost exactly the same output, just different format for underlying numbers

```{r}
anova(model1)
```


Calcualte F
```{r}
treatment.MS <- anova(model1)$"Mean Sq"[1]
resid.MS <- anova(model1)$"Mean Sq"[2]

treatment.MS/resid.MS
```





# And you might see this sometiems to

* aov() command instead of lm()
* this has a few different features that are useful down the road

```{r}
#use aov() to fit model
m.aov <- aov(fluoro ~ treatment, 
             data = dat)

summary(m.aov)
```



# What does this p-value mean?

* All of these p values are the same
* They mean that at least mean value of the 3 treatments is diffrent from the other
* But doesn't say which one!
* We might right in a paper "There were significant differences among treatments (F = 11.23, p = 0.0002)"