---
title: "Regression Diagnostics"
author: "brouwern@gmail.com"
date: "January 30, 2017"
output:
  word_document:
    reference_docx: rmarkdown_template.docx
    fig_width: 9
    fig_height: 5
    fig_caption: true
---

# Residual analysis: assessing assumptions of models

```{r, echo=FALSE}
source("./scripts_Lecture4/sx_reg_flowchart2_fit_diag_inf.R")
```


```{r, echo = F}

#load data
milk2 <- read.csv(file = "Skibiel_clean_milk_focal_column.csv")
milk3 <- read.csv(file = "Skibiel_clean_milk_focal_genera.csv")

# re-fit null model
lm.null <- lm(fat.percent ~ 1,
              data = milk3)
# re-fit female mass model
lm.mass <- lm(fat.percent ~ log(mass.female), data = milk3)

#gestation duration
lm.gestation <- lm(fat.percent ~ log(gest.month.NUM), data = milk3)

#lactation duration
lm.lac.duration <- lm(fat.percent ~ log(lacat.mo.NUM), data = milk3)

#Litter mass
lm.litter.mass <- lm(fat.percent ~ log(mass.litter), data = milk3)
```



# Residual analysis: assessing assumptions of models

## Assumptions of regression

* Normality
* Heterogeneity of variance
* Random sampling

# What are residuals?


```{r, echo = F}
par(mfrow = c(1,1),mar = c(4.4,4.4,2,1))
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk3,
     ylab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "Fat vs lactation duration")
abline(lm.lac.duration, col = 3, lwd = 3)
```



```{r, echo = F}
#Function to plot residuals as deviations from regression line
plot.my.residuals <- function(model,lwd.=2){
  y.hat <- predict(model)
  dat <- model$model
  segments(x0 = dat[,2], 
           x1 = dat[,2],
           y1 = dat[,1], 
           y0 = y.hat, col = 2, lty = 2,lwd = lwd.)
}

```


# Plot residuals

* aka e[i], errors
* Residuals are distance from the regression line to the observed data
* Regression line represents predictions from the regression equation
* Residual = predictions - observation
* Typically squared when doing any math
    + hence "squared errors" 

```{r, echo = F}
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk3,
     #ylab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "Fat vs lactation duration")
abline(lm.lac.duration, col = 3, lwd = 3)
plot.my.residuals(lm.lac.duration)
legend("topright",legend = "Residuals",col = 2,lty = 2)
```


## Compare residuals from Null and Alt hypothesis

```{r, echo = F}
par(mfrow = c(1,2))
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk3,
     #ylab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "Fat vs lactation duration")
abline(lm.null, col = 4, lwd = 3)
resid.Ho <- resid(lm.null)
y.hat.Ho <- predict(lm.null)
y <- milk3$fat.percent
x <- log(milk3$lacat.mo.NUM)
segments(x0 = x,x1 = x,y0 = y.hat.Ho,y1 = y, col     = 2, lty = 2,lwd = 2)

plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk3,
     #ylab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "Fat vs lactation duration")
abline(lm.lac.duration, col = 3, lwd = 3)
plot.my.residuals(lm.lac.duration)
legend("topright",legend = "Residuals",col = 2,lty = 2)
```


# What are model predictions?

* aka "y.hat"
* aka "fitted values"

```{r, echo = F}
par(mfrow = c(1,1),
    mar = c(5,5,3,2))
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk3,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "Predicted values")
abline(lm.lac.duration, col = 3, lwd = 3)
points(x = 0.833, y = 19, col = 3, pch = 19)
points(x = 0.833, y = 9.37, col = 4,cex = 1.3, pch = 19)

int <- round(coef(lm.lac.duration)[1],2)
slope <- round(coef(lm.lac.duration)[2],2)
eq <- paste("fat.percent ~ ",slope,
            " * log(lactation duration)",
            " + ",int, sep = "")
mtext(text = eq,side = 3,adj = 0)

eq2 <- paste("y.hat = ",slope,
             " * 0.833",
             " + ",int, sep = "")
text(x = 0.85, y = 16,
     label = eq2,pos = 4, cex = 1.5)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Diagnostics by hand


## Normality of residuals

* resids()

```{r, echo =F}
resids <- resid(lm.lac.duration)
hist(resids)
abline(v = 0,lwd = 4,lty = 2,col =3)
```

Note: residuals sum to approximatley zero (or very very very very very close)
```{r}
sum(resids)
```


## Heterogen of variance

* fitted() extracts predictions / fitted values

```{r}
fits <- fitted(lm.lac.duration)
plot(resids ~ fits)
abline(h = 1, lty = 3)
```
* What does this imply?


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



# Easy diagnostics

```{r}
par(mfrow = c(2,2), mar = c(5,5,2,2))
plot(lm.lac.duration)
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Diagnostic 1: Residuals vs. fitted

* Heterogen of variance
* Very important assumption!
* "which = 1" selects the resid vs. fitted plot

```{r}
par(mfrow = c(1,1))
plot(lm.lac.duration, which = 1)
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Diagnostic 2

* Assesses normality
* Not as important as heterogen of var
* (many reviews think its important)
* qqplot = "quantile-quantile plot"
* Trying to decide how much to explain this...
* Code: plot(model, which = 2)

```{r, echo = F}
par(mfrow = c(1,1))
plot(lm.lac.duration, which = 2)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Diagnostic 3

* Scale-location plot
* ...

```{r, echo = F}
par(mfrow = c(1,1))
plot(lm.lac.duration, which = 3)
```


## Diagnostic 4

* Outliers
* High leverage values have high impact on regression
* plot(model, which = 5)
* See handout

```{r, echo = F}
par(mfrow = c(1,1))
plot(lm.lac.duration, which = 5)
```



<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>





# Addressing non-constant variance (heteroskedasticity)

* Non-constant variance is a big issue
* can "model the variance" using gls()
* (generalized least squares)
* (note use of "na.omit")


```{r, echo = -1}
library(nlme)

milk.gls <- gls(fat.percent ~ log(lacat.mo.NUM), 
                data = milk3,
                na.action = na.omit,
                weights  = varPower())

summary(milk.gls)

```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Compare lm() and gls() models

* Does modeling variance improve model fit?
* our lm and gls models are NOT nested
* test w/ AIC

```{r, echo = -1}
library(bbmle)
AICtab(lm.lac.duration,
       milk.gls)
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Does this fix heterogen of var?

* Are residuals better?
* Kinda

```{r}
plot(milk.gls, pch = 18)
```



