---
title: "The t-test as regression"
author: "brouwern@gmail.com"
date: "January 26, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      cache = TRUE, warnings = FALSE, strip.white = F)
```



# The Data

Today we will work with the rat bladder data again.

Motulsky 2nd Ed, Chapter 30, page 220, Table 30.1.  Maximal relaxaction of muscle strips of old and young rat bladders stimualted w/ high concentrations of nonrepinephrine (Frazier et al 2006).  Response variable is %E.max

```{r}
#make dataframe with data in 2 sep. columns
df.rats <- data.frame(
old = c(20.8,2.8,50.0,
        33.3,29.4,38.9, 29.4,52.6,14.3),
young = c(45.5,55.0, 60.7, 
          61.5, 61.1, 65.5,42.9,
          37.5, NA))


```



# Reshape data

```{r, echo = F, cache = T,warning=FALSE}
#make stacked dataframe
library(reshape2)
df.rat.stack <- melt(df.rats,
                     variable.name = "rat.age",
                     value.name = "percent.E.max")


```

# Plot data

* You can put 2 plots next to each other using "par(mfrow = c(1,2)"
* A 2 x 2 grid would be "c(2,2)"
* and so on

```{r}
par(mfrow = c(1,2))
library(beeswarm)

#beeswarm plot of raw dta
beeswarm(percent.E.max ~ rat.age, 
         col = c(1,2),
         df.rat.stack)

#boxplot
boxplot(percent.E.max ~ rat.age, 
         #col = c(1,2),
         df.rat.stack)

```



# Plot means

* If you want just one plot again you have to reset par to "par(mfrow = c(1,1))"
* Note use of with()
```{r}
par(mfrow = c(1,1))
library(sciplot)
with(df.rat.stack,
     lineplot.CI(x.factor = rat.age,
                 response = percent.E.max,
                 #ylim = ylim.,
                 xlab = "Rat age",
     ylab = ""),
     )
```


# t-test


```{r}
t.test(percent.E.max ~rat.age,
       df.rat.stack,
       var.equal = T)
```

* NB: set " var.equal = T"
* would not normally do this
* t = 3.5315
* df = 15
* p = 0.003022


# t-test as linear regression

## Fit a linear regression

* linear regression = linear model = lm()
* assign output of model to an object
```{r}
rat.lm1 <- lm(percent.E.max ~ rat.age,
       df.rat.stack)
```

## Look at output of lm()

* To look at output of a linear regression use summary()
```{r}
summary(rat.lm1)
```



# t-test as a stat. model 

## Null hypothesis: No difference between old and young rats

* We can  represent this hypothesis as a model
* Instead of "~rat.age" we do "~1"
* So our equation is "percent.E.max ~ 1"
* This means "model percent.E.max just as a single mean value"

### Make model

```{r}
rat.lm.null <- lm(percent.E.max ~ 1,
       data = df.rat.stack)
```

### Look at Null model

```{r}
summary(rat.lm.null)
```


<br><br>

## Alt hypothesis:  difference between old & young rats

* We can  represent this hypothesis as a model
* So our equation is "percent.E.max ~ rat.age"
* This means "model percent.E.max with seperate means for each age group"
* (This is the same model we made above)

```{r}
rat.lm.alt <- lm(percent.E.max ~ rat.age,
       data = df.rat.stack)
```


### Look at Alt model

```{r}
summary(rat.lm.alt)
```


<br><br>

## Compare the models

One way of thinking about this hypothesis is, "If we wanted to predict E.max for  young rat in a new experiment, what's the best prediction?  The mean of all the rats from the 1st study, or the mean of the young rats?"

```{r}
anova(rat.lm.alt, rat.lm.null)
```




```{r}

AIC(rat.lm.null)
AIC(rat.lm.alt)

```






# t-test w/o versus w/ equal variance

What we just ran
```{r}
t.test(percent.E.max ~rat.age,
       df.rat.stack,
       var.equal = T)
```


The default is "var.equal = F"; don't have to include it
```{r}
t.test(percent.E.max ~rat.age,
       df.rat.stack,
       var.equal = F)
```



# linear regression correcting for unequal variance

* We can correct for unequal variance using the gls() function in the nlme package
* gls = "generalized least squares"
* nlme = "non-linear mixed effects"
* If you've done regression by hand in Excel, "least squares" is probably what you did

```{r}
library(nlme)

rat.gls <- gls(percent.E.max ~rat.age,
       df.rat.stack, 
       na.action = na.omit,
       weights  = varPower())
```



```{r}
summary(rat.gls)
```

