---
title: "Cleaning data"
author: "brouwern@gmail.com"
date: "April 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 

```{r}
list.files()
dat1 <- read.csv("grogan_bad_data.csv")


```


```{r}
dim(dat1)
```


# Extra columns to right of data

```{r}
head(dat1,15)
```


Can delete mannually in Excel or remove using indexing "negative indexing".

```{r}
dat1 <- dat1[,-c(5,6,7)]
```

Check

```{r}
head(dat1)
```


# Check factors

"fraction" and "abs" columns have issues
```{r}
summary(dat1)
```


# Fix bad factor with which()


## Fix typos

Fix the lower case "l" by changing to "L"
```{r}
i.l <- which(dat1$fraction == "l")

dat1$fraction[i.l]

dat1$fraction[i.l] <- "L"

```


```{r}
i.p <- which(dat1$fraction == "p")
i.s <- which(dat1$fraction == "s")


dat1$fraction[i.p] <- "P"
dat1$fraction[i.s] <- "S"
```

Could also do this w/ ifelse(), but I find that this function's behavior can do unexpected things.


## Check output

Note that the "l" and other lowercase factor names are still there!

```{r}
summary(dat1)
```


## Get ride of old factor levels


Change to character data
```{r}
dat1$fraction <- as.character(dat1$fraction)
```

Check
```{r}
summary(dat1)
```


Change back to factor
```{r}
dat1$fraction <- as.factor(dat1$fraction)
```


Check
```{r}
summary(dat1$fraction)
```


# Edit factor levels

## Change order of factor levels

Deafult is alphabetial.  Can use factor() function to reorder
```{r}
dat1$fraction <- factor(dat1$fraction,
                        levels = c("P","L","S"))
```


There is a function relevel() that can do somethign similar; however, it only sets the base or reference level and so isn't good if you have 3+ levels to the factor variable.

## Rename factor

```{r}
dat1$fraction <- factor(dat1$fraction,
                        labels =c("Px","Ly","Sz"))

```


```{r}
summary(dat1$fraction )
```





# Fix numeric data

```{r}
summary(dat1)
```


```{r}
dat1$abs
```


## Find the bad value using its exact label
```{r}
i.bad <- which(dat1$abs == "0.127069 a")

dat1$abs[i.bad]
```


## Find the bad value using a "Regular expression"

```{r}
i.bad.grep <- grep("a",dat1$abs)
```


```{r}
dat1$abs[i.bad.grep]
```


## Overwrite by hand

```{r}
dat1$abs[i.bad.grep] <- "0.127069"
```

Doesn't work!  Why?!?!!


## Convert to character data first
```{r}
dat1$abs <- as.character(dat1$abs)
```


```{r}
summary(dat1$abs)
```




## Make the change using index

```{r}
dat1$abs[i.bad.grep] <- "0.127069"
```

Conver to numeric

```{r}
dat1$abs <- as.numeric(dat1$abs)
```


Check

```{r}
summary(dat1$abs)
```


## Make the change using regular expression

Reload data
```{r}
dat2 <- read.csv("grogan_bad_data.csv")
```


```{r}
dat2$abs <- gsub("a","",dat1$abs)
```

No error!  Data converted to character data on the fly

```{r}
summary(dat2$abs)
```



```{r}
dat2$abs <- as.numeric(dat2$abs)
```

```{r}
summary(dat1$abs)
summary(dat2$abs)
```



# Reshape data

Data is in "long format"

```{r}
head(dat1)
```





```{r}
library(multcomp)

which(dat1$strain ==)

summary(glht(aov, linfct=mcp(Group="Dunnett")))
```


